**Urban mobility and socioeconomic deprivation in Latin America after COVID-19**

### 2024-10-26 manuscript/manuscript-nature/manuscript-nature.qmd
results1-fig1

---
title: "**Urban mobility and socioeconomic deprivation in Latin America after COVID-19**"
author:
  - name: Carmen Cabrera-Arnau
    affil-id: 1
  - name: Francisco Rowe
    affil-id: 1
  - name: Miguel González-Leonardo
    affil-id: 2
  - name: Andrea Nasuto
    affil-id: 1
  - name: Ruth Neville
    affil-id: 1
affiliations: 
  - id: 1
    name: Geographic Data Science Lab, Department of Geography and Planning, University of Liverpool, Liverpool, UK
  - id: 2
    name: Centre for Demographic Urban and Environmental Studies, El Colegio de México, Ciudad de México, México
abstract: |
  COVID-19 triggered a reduction in the frequency and extent of people's movement. Existing evidence suggests that while the impact of the pandemic on mobility was widespread, the extent of this impact was unequally felt across socioeconomic groups in the early stages of the pandemic. Here, we find that the most deprived locations have experienced a more accelerated recovery towards pre-pandemic levels of mobility in the long term. Furthermore, the socioeconomic disparities in the patterns of mobility triggered by the first outbreak of COVID-19, have persisted as of April 2023. These findings are based on the analysis of time-series mobility data corresponding to X urban areas from Latin American countries collected from Meta-Facebook users upon their consent. Our research highlights the importance of timely mobility data with high spatiotemporal resolution to understand the long-term effects of the pandemic and to inform equitable policy responses that address societal challenges in urban areas.
format: 
  pdf:
    fontsize: 11pt
    geometry:
      - left=2.54cm
      - right=2.54cm
    header-includes:
      \usepackage{lipsum}
      \usepackage{libertine}
    keep-tex: true
    template-partials: 
      - title.tex
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
number-sections: true
editor: visual
csl: nature.csl
bibliography: refs.bib
editor_options:
  markdown:
    wrap: sentence
---

<!--# FR: A few questions to think about: Have we validated the data? i.e. tested their extent of representativeness? Shall we do something quick using census data for each country comparing overall populations?  -->

\

# Introduction {#sec-intro}

@@

However, reductions in mobility levels were highly unequal reflecting existing socioeconomic inequalities in our societies [@chang2020].
In most countries, affluent individuals tended to record the greatest drops in mobility levels as they are predominantly employed in knowledge-intensive jobs which can be done fully or partly remotely [@fraiberger2020; @bonaccorsi2020; @weill2020; @dueñas2021; @santana2023].
During the COVID-19 pandemic, the adoption of remote work reduced the need of commuting for knowledge-intensive, non-public facing jobs [@florida2021].
At the same time, individuals from less privileged socioeconomic backgrounds displayed less pronounced declines mirroring the nature of their work requiring public-facing, face-to-face interaction, and thus a requirement for daily work commutes [@dueñas2021; @santana2023].

@@

# Results {#sec-results}

@@

The focus of this work is on capturing how mobility patterns for different population groups have been impacted by COVID-19 in Latin America.
For this purpose, we consider populations groups by classifying their local areas into categories according to their population density and their socioeconomic level.
These two criteria for classification are chosen due to their relevance for ...

We use five population density categories and three socioeconomic categories defined according to the relative deprivation index.
These categories are used to classify the spatial units of analysis, following the methodology defined in section X.
The geographic distribution of categories for each of the countries is displayed in Fig X.a), Fig X.a) and Fig X.a) for Argentina, Chile and Colombia respectively.
Fig X.b), Fig X.b) and Fig.b) show, for each country, the population distribution into the categories of analysis, considering, on the one hand, Worldpop population estimations and on the other hand, Facebook population, based on user counts during the pre-pandemic baseline period (see section X of the methodology for further details).
For the three countries we observe discrepancies in the population distribution according to WorldPop and Facebook data, which reflect the extent to which the Facebook data tends to overrepresent certain population groups with respect to the Worldpop data whilst underrepresenting others.
For example, for Argentina and Chile, the proportion of population living in areas of low population density is higher according to Worldpop data, hence indicating that Facebook data underrepresents populations in these regions.
Similarly, in those countries, Facebook data shows and overrepresentation of the most affluent socioeconomic group.

Biases in Facebook data arise for a variety of reasons.
EXPLAIN.
Finally, Meta shares Facebook data for active user population and flows that has been spatially and temporally aggregated.
Their data aggregation methodology is designed so that records corresponding to low counts are not included in the dataset in order to preserve the privacy of users.
Analysing aggregated Facebook data in its raw format can therefore result in an underrepresentation of groups that produce low data counts.
To counter these source of bias in the Facebook data, we propose a methodology to process the data, which is fully described in the section X.

The analysis of mobility is done using Facebook data that has been pre-processed according to the proposed methodology for bias correction.
Fig X.c), Fig X.c) and X.c) show, for Argentina, Chile and Colombia respectively, changes in the intensity of movement relative to pre-pandemic baseline levels, at two points during the period of analysis, May 2020 and March 2022.
The patterns are consistent across the three countries in both dimensions of I'M HERE!!!!!

@@

The evolution of the percentage change in the number of movements is measured with respect to a baseline period prior to the pandemic as described in @sec-methods.
For the purposes of the analysis, we aggregate the raw movement data temporally into months and spatially into administrative units at various levels according to GADM, the Database of Global Administrative Areas [@GADM24].
The analysis focuses on administrative units that are within the boundaries of functional urban areas as specified by the Global Human Settlement Layer (GHSL).
For each administrative unit, we compute the Relative Deprivation Index (RDI) based on data from NASA's Socioeconomic Data and Applications Centre (SEDAC).
Figure 1 displays the administrative units included in the study, coloured according to their average RDI.
Predictions about the evolution of the percentage change in the number of movements are made using the Prophet forecasting procedure.
Further details are provided in @sec-methods.

@@

# Data and Methods {#sec-methods}

@@

 !-->
```
# References

>>>>

---
title: Data Descriptor for Nature Scientific Data
author: 
## listed alphabetically
    - name: Carmen Cabrera-Arnau
      affiliation: 
        name: "1,* "
    - name: Francisco Rowe
      affiliation: 
        name: "1"
    - name: Miguel González-Leonardo
      affiliation: 
        name: "2"
    - name: Andrea Nasuto
      affiliation: 
        name: "1"
    - name: Ruth Neville
      affiliation: 
        name: "1"
authornotes:
    - note: "1"
      text: Geographic Data Science Lab, Department of Geography and Planning, University of Liverpool, Liverpool, UK
    - note: "2"
      text: Centre for Demographic Urban and Environmental Studies, El Colegio de México, Ciudad de México, México
    - note: "*"
      text: "`c.cabrera-arnau@liverpool.ac.uk`"
format:
  scientific-data-pdf: default
date: last-modified
abstract: |
  COVID-19 triggered a reduction in the frequency and extent of people's movement. Existing evidence suggests that while the impact of the pandemic on mobility was widespread, the extent of this impact was unequally felt across socioeconomic groups in the early stages of the pandemic. Here, we find that the most deprived locations have experienced a more accelerated recovery towards pre-pandemic levels of mobility in the long term. Furthermore, the socioeconomic disparities in the patterns of mobility triggered by the first outbreak of COVID-19, have persisted as of April 2023. These findings are based on the analysis of time-series mobility data corresponding to X urban areas from Latin American countries collected from Meta-Facebook users upon their consent. Our research highlights the importance of timely mobility data with high spatiotemporal resolution to understand the long-term effects of the pandemic and to inform equitable policy responses that address societal challenges in urban areas.

bibliography: bibliography.bib
---

# Main {.unnumbered}

<!--# CCA: Fr to rewrite section -->

@@

However, reductions in mobility levels were highly unequal reflecting existing socioeconomic inequalities in our societies [@chang2020].
In most countries, affluent individuals tended to record the greatest drops in mobility levels as they are predominantly employed in knowledge-intensive jobs which can be done fully or partly remotely [@fraiberger2020; @bonaccorsi2020; @weill2020; @dueñas2021; @santana2023].
During the COVID-19 pandemic, the adoption of remote work reduced the need of commuting for knowledge-intensive, non-public facing jobs [@florida2021].At the same time, individuals from less privileged socioeconomic backgrounds displayed less pronounced declines mirroring the nature of their work requiring public-facing, face-to-face interaction, and thus a requirement for daily work commutes [@dueñas2021; @santana2023].

@@

# Results

@@

We focus on capturing how mobility patterns for different population groups have been impacted by COVID-19 in Latin America.
We consider populations groups by classifying the spatial units of analysis into categories according to their population density and their socioeconomic level.
More details for the classification method are provided in section X.
The two criteria for classification are chosen due to their relevance for ...
We use five population density categories and three socioeconomic categories defined according to the relative deprivation index.
The geographic distribution of categories for each of the countries is displayed in Figures @Fig1-ARG a), @Fig1-CHL a) and @Fig1-COL a) for Argentina, Chile and Colombia respectively.

A key challenge in the in the analysis of population counts and movements with Facebook user data is the absence of small count records, which is a result of privacy-protection techniques applied to ensure that the location of individuals or small groups cannot be identified.
As a consequence, the countries of the analysis have locations with null values for Facebook population counts and flows.
These missing data are not distributed randomly.
For example, the missing values for FB population counts display high spatial autocorrelation as shown in Supplementary Figure SF1.
Therefore, simply removing the missing records from the analysis could lead to geographically biased results (Afghari et al., 2019).
To address this, we designed a data processing method to estimate the missing values, both for Facebook population counts and flows.
Furthermore, this method also applies a correction factor to eliminate the fluctuations in the daily number of observations, assuming that the representativeness of Facebook data across spatial units remains stable during the study period.
The data processing method is described in section X.

Even after applying the imputation method to estimate missing values, we acknowledge that Facebook data may still overrepresent certain population groups while underrepresenting others.
For example, Figures @Fig1-ARG b), @Fig1-CHL b) and @Fig1-COL b) show, for each country, the proportion of the population across the various analysis categories.
These comparisons use WorldPop population estimates and Facebook population counts, with the latter reflecting active users on an average day during a pre-pandemic baseline period (see Section X of the methodology for details).
In all three countries, notable discrepancies appear between the population distributions according to WorldPop and Facebook data.
For instance, in Argentina and Chile, WorldPop data indicates a higher proportion of people living in low-density areas, suggesting that Facebook data underrepresents populations in these regions.
Similarly, in these countries, Facebook data shows an overrepresentation of the most affluent socioeconomic group.
While addressing this kind of representativity bias is beyond the scope of this paper, we recognise its significance and the importance of addressing it in future analyses.
This issue is revisited in the Discussion section.

![](figures/Fig1.png)

The analysis of mobility is therefore done using Facebook movement data that has been pre-processed according to our methodology.
Figures @Fig1-ARG c), @Fig1-CHL c) and @Fig1-COL c) show, for Argentina, Chile and Colombia respectively, changes in the intensity of movement relative to pre-pandemic baseline levels, at two points during the period of analysis, May 2020 and March 2022.
The patterns show remarkable consistency across countries for both classification criteria.

@@

The evolution of the percentage change in the number of movements is measured with respect to a baseline period prior to the pandemic as described in Methods.
For the purposes of the analysis, we aggregate the raw movement data temporally into months and spatially into administrative units at various levels according to GADM, the Database of Global Administrative Areas [@GADM24].
The analysis focuses on administrative units that are within the boundaries of functional urban areas as specified by the Global Human Settlement Layer (GHSL).
For each administrative unit, we compute the Relative Deprivation Index (RDI) based on data from NASA's Socioeconomic Data and Applications Centre (SEDAC).
Figure 1 displays the administrative units included in the study, coloured according to their average RDI.
Predictions about the evolution of the percentage change in the number of movements are made using the Prophet forecasting procedure.
Further details are provided in Methods.

@@

# Data and methods {.unnumbered}

@@

!-->
```
# Code availability {.unnumbered}

For all studies using custom code in the generation or processing of datasets, a statement must be included under the heading "Code availability", indicating whether and how the code can be accessed, including any restrictions to access.
This section should also include information on the versions of any software used, if relevant, and any specific variables or parameters used to generate, test, or process the current dataset.

# References

::: {#refs}
:::

# Acknowledgements {.unnumbered}

The work was supported by Research England..., project no. ...
, ‘RECAST’.

# Author information {.unnumbered}

## Authors and affiliations

## Contributions

A.A.
conceived the experiment(s), A.A.
and B.A.
conducted the experiment(s), C.A.
and D.A.
analysed the results.
All authors reviewed the manuscript.

# Ethics declarations {.unnumbered}

## Competing interests

The authors declare no competing interests.

# Supplementary information {.unnumbered}

# Rights and permissions {.unnumbered}

Springer Nature or its licensor (e.g. a society or other partner) holds exclusive rights to this article under a publishing agreement with the author(s) or other rightsholder(s); author self-archiving of the accepted manuscript version of this article is solely governed by the terms of such publishing agreement and applicable law.

### 2024-10-26 manuscript/manuscript/manuscript.qmd
results1-fig1

---
title: "**Urban mobility and socioeconomic deprivation in Latin America after COVID-19**"
author:
  - name: Carmen Cabrera-Arnau
    affil-id: 1
  - name: Francisco Rowe
    affil-id: 1
  - name: Miguel González-Leonardo
    affil-id: 2
  - name: Andrea Nasuto
    affil-id: 1
  - name: Ruth Neville
    affil-id: 1
affiliations: 
  - id: 1
    name: Geographic Data Science Lab, Department of Geography and Planning, University of Liverpool, Liverpool, UK
  - id: 2
    name: Centre for Demographic Urban and Environmental Studies, El Colegio de México, Ciudad de México, México
abstract: |
  COVID-19 triggered a reduction in the frequency and extent of people's movement. Existing evidence suggests that while the impact of the pandemic on mobility was widespread, the extent of this impact was unequally felt across socioeconomic groups in the early stages of the pandemic. Here, we find that the most deprived locations have experienced a more accelerated recovery towards pre-pandemic levels of mobility in the long term. Furthermore, the socioeconomic disparities in the patterns of mobility triggered by the first outbreak of COVID-19, have persisted as of April 2023. These findings are based on the analysis of time-series mobility data corresponding to X urban areas from Latin American countries collected from Meta-Facebook users upon their consent. Our research highlights the importance of timely mobility data with high spatiotemporal resolution to understand the long-term effects of the pandemic and to inform equitable policy responses that address societal challenges in urban areas.
format: 
  pdf:
    fontsize: 11pt
    geometry:
      - left=2.54cm
      - right=2.54cm
    header-includes:
      \usepackage{lipsum}
      \usepackage{libertine}
    keep-tex: true
    template-partials: 
      - title.tex
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
number-sections: true
editor: visual
csl: nature.csl
bibliography: refs.bib
editor_options:
  markdown:
    wrap: sentence
---

@@

The focus of this work is on capturing how mobility patterns for different population groups have been impacted by COVID-19 in Latin America.
For this purpose, we consider populations groups by classifying their local areas into categories according to their population density and their socioeconomic level.
These two criteria for classification are chosen due to their relevance for ...

We use five population density categories and three socioeconomic categories defined according to the relative deprivation index.
These categories are used to classify the spatial units of analysis, following the methodology defined in section X.
The geographic distribution of categories for each of the countries is displayed in Fig X.a), Fig X.a) and Fig X.a) for Argentina, Chile and Colombia respectively.
Fig X.b), Fig X.b) and Fig.b) show, for each country, the population distribution into the categories of analysis, considering, on the one hand, Worldpop population estimations and on the other hand, Facebook population, based on user counts during the pre-pandemic baseline period (see section X of the methodology for further details).
For the three countries we observe discrepancies in the population distribution according to WorldPop and Facebook data, which reflect the extent to which the Facebook data tends to overrepresent certain population groups with respect to the Worldpop data whilst underrepresenting others.
For example, for Argentina and Chile, the proportion of population living in areas of low population density is higher according to Worldpop data, hence indicating that Facebook data underrepresents populations in these regions.
Similarly, in those countries, Facebook data shows and overrepresentation of the most affluent socioeconomic group.

Biases in Facebook data arise for a variety of reasons.
EXPLAIN.
Finally, Meta shares Facebook data for active user population and flows that has been spatially and temporally aggregated.
Their data aggregation methodology is designed so that records corresponding to low counts are not included in the dataset in order to preserve the privacy of users.
Analysing aggregated Facebook data in its raw format can therefore result in an underrepresentation of groups that produce low data counts.
To counter these source of bias in the Facebook data, we propose a methodology to process the data, which is fully described in the section X.

The analysis of mobility is done using Facebook data that has been pre-processed according to the proposed methodology for bias correction.
Fig X.c), Fig X.c) and X.c) show, for Argentina, Chile and Colombia respectively, changes in the intensity of movement relative to pre-pandemic baseline levels, at two points during the period of analysis, May 2020 and March 2022.
The patterns are consistent across the three countries in both dimensions of I'M HERE!!!!!

>>>>

---
title: "**Urban mobility and socioeconomic deprivation in Latin America after COVID-19**"
author:
  - name: Carmen Cabrera-Arnau
    affil-id: 1
  - name: Francisco Rowe
    affil-id: 1
  - name: Miguel González-Leonardo
    affil-id: 2
  - name: Andrea Nasuto
    affil-id: 1
  - name: Ruth Neville
    affil-id: 1
affiliations: 
  - id: 1
    name: Geographic Data Science Lab, Department of Geography and Planning, University of Liverpool, Liverpool, UK
  - id: 2
    name: Centre for Demographic Urban and Environmental Studies, El Colegio de México, Ciudad de México, México
abstract: |
  COVID-19 triggered a reduction in the frequency and extent of people's movement. Existing evidence suggests that while the impact of the pandemic on mobility was widespread, the extent of this impact was unequally felt across socioeconomic groups in the early stages of the pandemic. Here, we find that the most deprived locations have experienced a more accelerated recovery towards pre-pandemic levels of mobility in the long term. Furthermore, the socioeconomic disparities in the patterns of mobility triggered by the first outbreak of COVID-19, have persisted as of April 2023. These findings are based on the analysis of time-series mobility data corresponding to X urban areas from Latin American countries collected from Meta-Facebook users upon their consent. Our research highlights the importance of timely mobility data with high spatiotemporal resolution to understand the long-term effects of the pandemic and to inform equitable policy responses that address societal challenges in urban areas.

format: 
  nature-pdf: default
  pdf:
    fontsize: 11pt
    geometry:
      - left=2.54cm
      - right=2.54cm
    header-includes:
      \usepackage{lipsum}
      \usepackage{libertine}
    keep-tex: true
    template-partials: 
      - title.tex
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
number-sections: true
editor: visual
csl: nature.csl
bibliography: refs.bib
editor_options:
  markdown:
    wrap: sentence
---

@@

The focus of this work is on capturing how mobility patterns for different population groups have been impacted by COVID-19 in Latin America.
For this purpose, we consider populations groups by classifying their local areas into categories according to their population density and their socioeconomic level.
These two criteria for classification are chosen due to their relevance for ...
We use five population density categories and three socioeconomic categories defined according to the relative deprivation index.
These categories are used to classify the spatial units of analysis, following the methodology defined in section X.
The geographic distribution of categories for each of the countries is displayed in Fig X.a), Fig X.a) and Fig X.a) for Argentina, Chile and Colombia respectively.

A key challenge in the in the analysis of population counts and movements with Facebook user data is the absence of small count records, which is a result of privacy-protection techniques applied to ensure that the location of individuals or small groups cannot be identified.
As a consequence, the countries of the analysis have locations with null values for population counts and flows.
Given that these missing data are not distributed randomly (Supplementary Table A1), simply deleting the incomplete records could lead to biased results (Afghari et al., 2019).
Therefore, we designed an imputation method to estimate these missing values.
The overview of this missing data imputation for Facebook user count is presented in Fig.
1.

Even after applying the imputation method for the estimation of missing values, Facebook data tends to overrepresent certain population groups with respect to Wordlpop data whilst underrepresenting others.
Fig X.b), Fig X.b) and Fig.b) show, for each country, the share of population in each of the categories of analysis, considering, on the one hand, Worldpop population estimations and on the other hand, Facebook population, based on user counts during the pre-pandemic baseline period (see section X of the methodology for further details).
For the three countries we observe discrepancies in the population distribution according to WorldPop and Facebook data.
For example, for Argentina and Chile, the share of population living in areas of low population density is higher according to Worldpop data, hence indicating that Facebook data underrepresents populations in these regions.
Similarly, in those countries, Facebook data shows an overrepresentation of the most affluent socioeconomic group.
"Addressing this type of representativity bias is beyond the scope of this paper, but we are aware of it and we recognise the importance of addressing this in future analysis".

The analysis of mobility is therefore done using Facebook movementdata that has been pre-processed for imputation of low counts.
Fig X.c), Fig X.c) and X.c) show, for Argentina, Chile and Colombia respectively, changes in the intensity of movement relative to pre-pandemic baseline levels, at two points during the period of analysis, May 2020 and March 2022.
The patterns are consistent across the three countries in both dimensions of I'M HERE!!!!!
